Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-20 submission/manuscript.Rmd
edit discussion p3

While many genes involved in early germline development are common between males and females, some display sex-biased expression or perform functions unique to eggs and sperm. We found both males and females expressed both sperm and egg biased genes, indicating the mechanisms underlying germline gene repression are independent of germ cell sex. However, we found organismal sex did greatly influence the manifestation of germline gene dysregulation, as _Kdm5c_-KO females had over double the number of germline-enriched DEGs compared to male EpiLCs. Knockout of KDM5C in females is embryonic lethal and results in impaired X chromosome inactivation (XCI) [@samantaActivationXistEvolutionarily2022]. XCI defects could explain why homozygous _Kdm5c_ knockout females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, ectopic germline transcripts, including those unique to females, were not significantly biased towards the X chromsome. Sex differences in germline gene suppression may be instead connected to females having a higher dose of KDM5C than males, due to its partial escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, heterozygous loss of _Kdm5c_ in females also had over double the number of germline DEGs than males, even though their expression of KDM5C should be roughly equivalent to that of wild-type males. Altogether, these results suggest female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity. Future studies investigating sex differences in germline gene repression are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators or a general feature of female cells.

>>>>

Although eggs and sperm intially express very similar sets of germline genes for PGC formation and meiosis, some germline genes are sexually dimorphic. We found _Kdm5c_ mutant males and females expressed both sperm and egg-biased genes, indicating the mechanism of repression is independent of a given germline gene's sex. However, organismal sex did greatly influence the manifestation of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males. X chromosome inactivation (XCI)  defects could explain why homozygous _Kdm5c_ knockout females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004] and loss of _Kdm5c_ impairs XCI[@samantaActivationXistEvolutionarily2022]. However, female germline DEGs were not significantly biased towards the X chromosome. Sex differences in germline gene suppression may be instead connected to females having a higher dose of KDM5C than males, due to its partial escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than males, even though their level of KDM5C should be roughly equivalent to that of wild-type males. Altogether, these results suggest female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity. Future studies are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators or a general feature of female cells.
